WASP-4\lowercase{b} Arrived Early for the TESS Mission

### 2019-03-27 paper/ms.tex
caught bug in k2p reported errors
reported just the 84th and 16th percentiles in the arxiv paper, instead
of 84-50, and 50-16.

(not good!)
update k2p discussion

\begin{figure*}[t]
    \begin{center}
        \includegraphics[width=0.9\textwidth]{f2.png}
    \end{center}
    \vspace{-0.7cm}
    \caption{
        {\bf Phase-folded lightcurve of WASP-4b.} Gray points are
        TESS \replaced{data}{flux measurements, with median 
        $1\sigma$ uncertainty shown in the lower 
        right.} 
        Yellow points are binned measurements.  The bottom
        panel shows the residuals.  The fit to the phase-folded
        transit (blue line) is used when measuring mid-transit times
        \replaced{for the}{of} individual transits (see 
        \S~\ref{sec:measurement}).
        \label{fig:phasefold}
    }
\end{figure*}

@@

\subsection{Analysis}

@@

\added{We fitted each model by assuming a Gaussian likelihood and
sampling over the posterior probability distributions. The prior for
the quadratic model allowed the period derivative to have any sign.}
Figure~\ref{fig:times} shows the residuals with respect to the
constant-period model.  The best-fitting constant-period model has
$\chi^2 = 174$ and 61 degrees of freedom.  The best-fitting
\replaced{decreasing-period}{quadratic} model has $\chi^2 = 62.6$ and
60 degrees of freedom.  The best-fitting precession model has $\chi^2
= 64.3$ and 58 degrees of freedom.  Either of the latter two models
provides a much better fit than the constant-period model.  The
difference in $\chi^2$ between the first and second models corresponds
to $p \approx 10^{-26}$.

The \replaced{decreasing-period}{quadratic} model provides a slightly
better fit to the data than the precession model.  It is favored by
$\Delta \chi^2 = 1.7$, and has two fewer free parameters.  A useful
heuristic for model comparison is the Bayesian Information Criterion
(BIC),
\begin{equation}
  {\rm BIC} = \chi^2 + k\log n,
\end{equation}
where $k$ is the number of free parameters, and $n$ is the number of
data points. In this case, $n=62$.  The difference in the BIC between
the precession and decay models is $\Delta {\rm BIC} = {\rm BIC}_{\rm
prec} - {\rm BIC}_{\rm quad} = 10$, corresponding to a Bayes factor of
$1.1\times 10^{4}$.  Likewise, the Akaike Information Criterion favors
the constant-period-derivative model by $\Delta {\rm AIC} = 5.8$.
Differences of this magnitude are traditionally deemed ``strong
evidence'' that one model is a better description of the data than the
other \citep{kass_bayes_1995}, although we prefer to reserve judgment
until more data can be obtained.

@@

To assess the robustness of these results, we considered a few
possible systematic effects in the timing dataset.

@@

With spots injected, we repeated our measurement of the transit times.
On average, the measured transit times did not change after injecting
spots, because the flux deviations are equally likely to occur in the
first and second halves of the transit.  For individual transits,
there were no cases for which the timing deviation was larger than one
minute.  The largest shifts occur when the spot anomaly occurs during
transit ingress or egress, in which case the measured mid-time is
shifted either late or early by between 30 and 50 seconds
\citep[qualitatively similar to results found
by][]{ioannidis_how_2016}.

Therefore the TESS observations could hypothetically all be skewed
early if there were spot-crossing events during every egress.  Two
arguments rule out this possibility.  (1) The lightcurve residuals do
not show evidence for these events.  (2) The stellar rotation period
is between 20 and 40 days, and the sky-projected stellar obliquity is
less than 10 degrees
\citep{triaud_spin-orbit_2010,sanchis-ojeda_starspots_2011,hoyer_tramos_2013}.
Since the planet orbits every 1.3 days, requiring that spot anomalies
always occur during egress would be equivalent to requiring a stellar
spot distribution that is exquisitely (and thus implausibly)
distributed to match the planet egress times.

@@

If instead the observed timing variation is just a small portion of an
apsidal precession cycle, then the orbital eccentricity is a few times
$10^{-3}$, and the full precession period is about 27 years.
\citet{ragozzine_probing_2009} calculated apsidal precession periods
for hot Jupiters, finding them to range between about 10 and 100
years. They highlighted that for many hot Jupiters, including WASP-4b,
the theoretical precession rate is dominated by the non-Keplerian
force due to the planet's tidal bulge.  Precession from general
relativity, the planet's rotational bulge, and the star's rotational
and tidal bulges contribute at the 10\% level at most.  Thus, a
measurement of the precession rate can be used to determine the
planet's Love number.  From their Equation 14, the implied Love number
for WASP-4b is
\begin{equation}
k_{2,{\rm p}} = 1.5^{+2.1}_{-1.1}.
\end{equation}
For comparison, the Love number of Jupiter is about 0.55
\citep{wahl_tidal_2016,ni_empirical_2018}, and a uniform density
sphere has $k_2 = 1.5$. The uncertainty in $k_{2,{\rm p}}$ for WASP-4b
is large because the eccentricity, reference time, and ${\rm
d}\omega/{\rm d}E$ have strongly correlated errors, and the measured
occultation times only barely constrain these parameters
(Figure~\ref{fig:future}).

The main problem with the apsidal precession hypothesis is to explain
why the eccentricity would be as large as $\sim$$10^{-3}$ despite
rapid tidal circularization.  For WASP-4, Equation~\ref{eq:de_dt}
gives $\tau_e = 0.29 (Q_{\rm p}'/10^5) \, {\rm Myr}$.  The star is
several billion years old, so unless the planet arrived very recently,
any initial eccentricity should have been lowered well below
$\sim$$10^{-3}$.  The top panel of Figure~\ref{fig:context} compares
the expected eccentricity damping time of WASP-4b with that of other
transiting giant planets.  WASP-4b has one of the shortest known
eccentricity damping times.

@@

\explain{The following subsection had material moved into the ``Timing
variation due to line of sight acceleration'' subsection.}
\subsection{Other possible explanations}

\begin{figure*}[t]
	\begin{center}
		\leavevmode
		\includegraphics[width=0.9\textwidth]{f7.pdf}
	\end{center}
	\vspace{-0.7cm}
	\caption{
    {\bf Further observations will be needed to confirm and understand
    the timing variations of WASP-4b.} \replaced{Dots}{Symbols} are as
    in Figure~\ref{fig:times}.  Lines are 100 random draws from the
    posteriors of the apsidal precession model (orange), and the
    orbital decay model (blue).  \added{The two models may begin to
    diverge in the mid-2020s}.    
		\label{fig:future}
	}
\end{figure*}

@@

If TESS is extended beyond its primary mission, it will likely observe
additional transits of WASP-4b in the early 2020s
(Figure~\ref{fig:future}).  High-precision transit observations with
larger telescopes would also be useful.  In order to decide between
orbital decay and apsidal precession, occultation measurements in both
the near term and also in the mid-2020s will be needed.  More
radial-velocity data would help in the search for additional bodies
that could be causing dynamical perturbations, or an overall
acceleration of the host star.  The transit duration variations are
expected to be of order 10 seconds \citep{pal_periastron_2008}, and so
may remain out of reach.

@@

\acknowledgements
L.G.B.\ and J.N.W.\ gladly acknowledge helpful discussions with
A.~Bailey, J.~Goodman, K.~Patra and V.~Van Eylen, and are
grateful to the people who have turned TESS from an idea into reality.
\added{L.G.B. thanks A.~Bixel and E.~May for clarifying details
concerning the available IMACS lightcurves.}
%
WASP-4 was included on the ``short-cadence'' target list thanks to the
Guest Investigator programs of J.\ Southworth and S.\ Kane (G011112
and G011183 respectively). 
%
J.N.W.\ thanks the TESS project and the Heising-Simons foundation for
supporting this work.
%
T.D. acknowledges support from MIT's Kavli Institute as a Kavli
postdoctoral fellow.
%
J.M.D. acknowledges funding from the European Research Council (ERC)
under the European Union's Horizon 2020 research and innovation
programme (grant agreement no. 679633; Exo-Atmos), and support from
the NWO TOP Grant Module 2 (Project Number 614.001.601).
%
J.E.R. was supported by the Harvard Future Faculty Leaders
Postdoctoral fellowship.
%
This paper includes data collected by the TESS mission, which are
publicly available from the Mikulski Archive for Space Telescopes
(MAST).
%
Funding for the TESS mission is provided by NASA's Science Mission
directorate.
%
This research has made use of the NASA Exoplanet Archive, which is
operated by the California Institute of Technology, under contract
with the National Aeronautics and Space Administration under the
Exoplanet Exploration Program.
%
This work made use of NASA's Astrophysics Data System Bibliographic
Services.
%
This research has made use of the VizieR catalogue access tool, CDS,
Strasbourg, France. The original description of the VizieR service was
published in A\&AS 143, 23.
%
This work has made use of data from the European Space Agency (ESA)
mission {\it Gaia} (\url{https://www.cosmos.esa.int/gaia}), processed
by the {\it Gaia} Data Processing and Analysis Consortium (DPAC,
\url{https://www.cosmos.esa.int/web/gaia/dpac/consortium}). Funding
for the DPAC has been provided by national institutions, in particular
the institutions participating in the {\it Gaia} Multilateral
Agreement.
%
\newline
%
\facility{
	TESS \citep{ricker_transiting_2015},
	Gaia \citep{gaia_collaboration_gaia_2016,gaia_collaboration_gaia_2018}
}
%
\software{
  \texttt{astrobase} \citep{bhatti_astrobase_2018},
  \texttt{astropy} \citep{the_astropy_collaboration_astropy_2018},
  \texttt{astroquery} \citep{astroquery_2018},
  \texttt{BATMAN} \citep{kreidberg_batman_2015},
  \texttt{corner} \citep{corner_2016},
  \texttt{emcee} \citep{foreman-mackey_emcee_2013},
  \texttt{IPython} \citep{perez_2007},
  \texttt{matplotlib} \citep{hunter_matplotlib_2007}, 
  \texttt{MESA} 
  \citep{paxton_modules_2011,paxton_modules_2013,paxton_modules_2015}
  \texttt{numpy} \citep{walt_numpy_2011}, 
  \texttt{pandas} \citep{mckinney-proc-scipy-2010},
  \texttt{radvel} \citep{fulton_radvel_2018},
  \texttt{scikit-learn} \citep{scikit-learn},
  \texttt{scipy} \citep{jones_scipy_2001}.
}

>>>>

\begin{figure*}[t]
    \begin{center}
        \includegraphics[width=0.7\textwidth]{f2.png}
    \end{center}
    \vspace{-0.7cm}
    \caption{
        {\bf Phase-folded lightcurve of WASP-4b.} Gray points are
        TESS \replaced{data}{flux measurements, with median 
        $1\sigma$ uncertainty shown in the lower 
        right.} 
        Yellow points are binned measurements.  The bottom
        panel shows the residuals.  The fit to the phase-folded
        transit (blue line) is used when measuring mid-transit times
        \replaced{for the}{of} individual transits (see 
        \S~\ref{sec:measurement}).
        \label{fig:phasefold}
    }
\end{figure*}

@@

\subsection{Analysis}
\label{subsec:analysis}

@@

\added{We fitted each model by assuming a Gaussian likelihood and
sampling over the posterior probability distributions. The prior for
the quadratic model allowed the period derivative to have any sign.
We considered two possible priors for the precession model: the first
is a wide prior that allows non-physical values of the planetary
Love number (Equation~\ref{eq:love_number}). The second requires
that the planetary Love number is less than that of a sphere with a
homogeneous density profile.  }

Figure~\ref{fig:times} shows the residuals with respect to the
constant-period model.  The best-fitting constant-period model has
$\chi^2 = 174$ and 61 degrees of freedom.  The best-fitting
\replaced{decreasing-period}{quadratic} model has $\chi^2 = 62.6$ and
60 degrees of freedom.  The best-fitting precession model \added{under
the wide prior} has \replaced{$\chi^2 = 64.3$}{$\chi^2 = 64.0$} and 58
degrees of freedom.\added{ Under the physical prior, the fit is slightly
worse, with $\chi^2=64.8$}. The precession and quadratic models both
provide much better fits than the constant-period model.  The
difference in $\chi^2$ between the linear and quadratic models
corresponds to $p \approx 10^{-26}$.

The \replaced{decreasing-period}{quadratic} model provides a slightly
better fit to the data than the precession model.  It is favored by
\replaced{$\Delta \chi^2 = 1.7$}{$\Delta \chi^2 = 1.4$}, and has two
fewer free parameters.  A useful heuristic for model comparison is the
Bayesian Information Criterion,
\begin{equation}
  {\rm BIC} = \chi^2 + k\log n,
\end{equation}
where $k$ is the number of free parameters, and $n$ is the number of
data points. In this case, $n=62$.  The difference in the BIC between
the precession and decay models is $\Delta {\rm BIC} = {\rm BIC}_{\rm
prec} - {\rm BIC}_{\rm quad} = 10$\added{, assuming a wide prior for
the precession model}. This corresponds to a Bayes factor of
$1.1\times 10^{4}$.  Likewise, the Akaike Information Criterion favors
the constant-period-derivative model by $\Delta {\rm AIC} = 5.8$.
Differences of this magnitude are traditionally deemed ``strong
evidence'' that one model is a better description of the data than the
other \citep{kass_bayes_1995}, although we prefer to reserve judgment
until more data can be obtained.

@@

To assess the overall robustness of our results, we considered a few
possible systematic effects in the timing dataset.

@@

With spots injected, we repeated our measurement of the transit times.
On average, the measured transit times did not change after injecting
spots, because the flux deviations are equally likely to occur in the
first and second halves of the transit.  For individual transits,
there were no cases for which the timing deviation was larger than one
minute.  The largest shifts occur when the spot anomaly occurs during
transit ingress or egress, in which case the measured mid-time is
shifted either late or early by between 30 and 50 seconds.

The TESS observations could therefore be skewed
early if there were spot-crossing events during {\it every} egress.  Two
arguments rule out this possibility.  (1) The lightcurve residuals do
not show evidence for these events.  (2) The stellar rotation period
is between 20 and 40 days, and the sky-projected stellar obliquity is
less than 10 degrees
\citep{triaud_spin-orbit_2010,sanchis-ojeda_starspots_2011,hoyer_tramos_2013}.
Since the planet orbits every 1.3 days, requiring that spot anomalies
always occur during egress would be equivalent to requiring a stellar
spot distribution that is exquisitely (and thus implausibly)
distributed to match the planet egress times.

@@

\explain{The result given in Equation~\ref{eq:love_number} has been
corrected from the submitted version. The previously stated result, k2
= $1.5^{+2.1}_{-1.1}$ had a whopping three errors. The first was a
typo that led to printing of 84th and 16th percentiles, instead of the
difference of these percentiles from the median. The second was an
insufficiently precise specification of the assumed priors, which are
now described.  The third was ignoring the uncertainty in the planet
radius and semimajor axis, which enter at the fifth power and are now
included.  }

If instead the observed timing variation is just a small portion of an
apsidal precession cycle, then the orbital eccentricity is a few times
$10^{-3}$, and the full precession period is about 27 years.
\citet{ragozzine_probing_2009} calculated apsidal precession periods
for hot Jupiters, finding them to range between about 10 and 100
years. They highlighted that for many hot Jupiters, including WASP-4b,
the theoretical precession rate is dominated by the non-Keplerian
force due to the planet's tidal bulge.  Precession from general
relativity, the planet's rotational bulge, and the star's rotational
and tidal bulges contribute at the 10\% level at most.  Thus, a
measurement of the precession rate can be used to determine the
planet's Love number.  From their Equation 14, the implied Love number
for WASP-4b is
\begin{align}
  k_{2,{\rm p}} &= 1.59^{+0.70}_{-0.47}
    \quad{\rm (wide\ prior),} \\
  k_{2,{\rm p}} &=  1.20^{+0.20}_{-0.26}
    \quad(k_{2,{\rm p}} \sim \mathcal{U}[0.015, 1.5])
  \label{eq:love_number}
\end{align}
where the uncertainties in the semimajor axis and planet radius have been
propagated through our Markov chains, and $\mathcal{U}$ denotes the uniform 
distribution.  For comparison, the Love number
of Jupiter is about 0.55 \citep{wahl_tidal_2016,ni_empirical_2018},
and a uniform density sphere has $k_2 = 1.5$. The uncertainty in
$k_{2,{\rm p}}$ for WASP-4b is large because the eccentricity,
reference time, and ${\rm d}\omega/{\rm d}E$ have strongly correlated
errors, and the \added{four available}\deleted{measured} occultation
times only weakly constrain these parameters
(Figure~\ref{fig:future}).

\added{ The results under both priors are larger than the Love number
of Jupiter, at weak statistical signifiance.  Imposing the
physically-motivated prior truncates the allowed values of $k_{2,{\rm
p}}$, and drives the precession period towards larger values.
Different priors also change the predicted evolution of the
orbit.  While the precession model matches the orbital decay model for
the next two decades in transits (Figure~\ref{fig:future}), in
occultations the precession model can deviate from the orbital decay
model, particularly if one imposes stricter priors on the planetary
Love number. This suggests that promptly obtaining occultation time
measurements for the system could help rule out the apsidal precession
model.  For the time being, the data are insufficient to make a
stronger ruling.  }

The main problem with the apsidal precession hypothesis \added{under
either of the priors} is to explain why the eccentricity would be as
large as $\sim$$10^{-3}$ despite rapid tidal circularization.  For
WASP-4, Equation~\ref{eq:de_dt} gives $\tau_e = 0.29 (Q_{\rm p}'/10^5)
\, {\rm Myr}$.  The star is several billion years old, so unless the
planet arrived very recently, any initial eccentricity should have
been lowered well below $\sim$$10^{-3}$.  The top panel of
Figure~\ref{fig:context} compares the expected eccentricity damping
time of WASP-4b with that of other transiting giant planets.  WASP-4b
has one of the shortest known eccentricity damping times.

@@

\explain{The following subsection had material moved into the ``Timing
variation due to line of sight acceleration'' subsection. Figure 7 was
updated to clarify the Love number prior for the precession model.}
\subsection{Other possible explanations}

\begin{figure*}[!ht]
	\begin{center}
		\leavevmode
		\includegraphics[width=0.9\textwidth]{f7.pdf}
	\end{center}
	\vspace{-0.7cm}
	\caption{
    {\bf Further observations are needed to confirm and understand
    the timing variations of WASP-4b.} \replaced{Dots}{Symbols} are as
    in Figure~\ref{fig:times}.  Lines are 100 random draws from the
    posteriors of the apsidal precession model (orange), and the
    orbital decay model (blue).  \added{The prior for the precession
    model assumes $k_{2,{\rm p}} \sim \mathcal{U}[0.015, 1.5]$; dashed orange
    lines emphasize samples with planetary Love numbers below 0.75}.    
		\label{fig:future}
	}
\end{figure*}

@@

If TESS is extended beyond its primary mission, it will likely observe
additional transits of WASP-4b in the early 2020s.
High-precision transit observations with
larger telescopes would also be useful.  In order to decide between
orbital decay and apsidal precession, occultation measurements in both
the near term and also in the mid-2020s will be needed 
(Figure~\ref{fig:future}).  More
radial-velocity data would help in the search for additional bodies
that could be causing dynamical perturbations, or an overall
acceleration of the host star.  The transit duration variations are
expected to be of order 10 seconds \citep{pal_periastron_2008}, and so
may remain out of reach.

@@

\acknowledgements
L.G.B.\ and J.N.W.\ gladly acknowledge helpful discussions with
A.~Bailey, J.~Goodman, K.~Patra\added{, D. Ragozzine,} and V.~Van
Eylen, and are grateful to the people who have turned TESS from an
idea into reality.  \added{L.G.B. thanks A.~Bixel and E.~May for
clarifying details concerning the available IMACS lightcurves.}
%
WASP-4 was included on the ``short-cadence'' target list thanks to the
Guest Investigator programs of J.\ Southworth and S.\ Kane (G011112
and G011183 respectively). 
%
J.N.W.\ thanks the TESS project and the Heising-Simons foundation for
supporting this work.
%
T.D. acknowledges support from MIT's Kavli Institute as a Kavli
postdoctoral fellow.
%
J.M.D. acknowledges funding from the European Research Council (ERC)
under the European Union's Horizon 2020 research and innovation
programme (grant agreement no. 679633; Exo-Atmos), and support from
the NWO TOP Grant Module 2 (Project Number 614.001.601).
%
J.E.R. was supported by the Harvard Future Faculty Leaders
Postdoctoral fellowship.
%
This paper includes data collected by the TESS mission, which are
publicly available from the Mikulski Archive for Space Telescopes
(MAST).
%
Funding for the TESS mission is provided by NASA's Science Mission
directorate.
%
This research has made use of the NASA Exoplanet Archive, which is
operated by the California Institute of Technology, under contract
with the National Aeronautics and Space Administration under the
Exoplanet Exploration Program.
%
This work made use of NASA's Astrophysics Data System Bibliographic
Services.
%
This research has made use of the VizieR catalogue access tool, CDS,
Strasbourg, France. The original description of the VizieR service was
published in A\&AS 143, 23.
%
This work has made use of data from the European Space Agency (ESA)
mission {\it Gaia} (\url{https://www.cosmos.esa.int/gaia}), processed
by the {\it Gaia} Data Processing and Analysis Consortium (DPAC,
\url{https://www.cosmos.esa.int/web/gaia/dpac/consortium}). Funding
for the DPAC has been provided by national institutions, in particular
the institutions participating in the {\it Gaia} Multilateral
Agreement.
%
\newline
%
\facility{
	TESS \citep{ricker_transiting_2015},
	Gaia \citep{gaia_collaboration_gaia_2016,gaia_collaboration_gaia_2018}
}
%
\software{
  \texttt{astrobase} \citep{bhatti_astrobase_2018},
  \texttt{astropy} \citep{the_astropy_collaboration_astropy_2018},
  \texttt{astroquery} \citep{astroquery_2018},
  \texttt{BATMAN} \citep{kreidberg_batman_2015},
  \texttt{corner} \citep{corner_2016},
  \texttt{emcee} \citep{foreman-mackey_emcee_2013},
  \texttt{IPython} \citep{perez_2007},
  \texttt{matplotlib} \citep{hunter_matplotlib_2007}, 
  \texttt{MESA} 
  \citep{paxton_modules_2011,paxton_modules_2013,paxton_modules_2015}
  \texttt{numpy} \citep{walt_numpy_2011}, 
  \texttt{pandas} \citep{mckinney-proc-scipy-2010},
  \texttt{radvel} \citep{fulton_radvel_2018},
  \texttt{scikit-learn} \citep{scikit-learn},
  \texttt{scipy} \citep{jones_scipy_2001}.
}
